\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{parskip}
\usepackage[colorlinks=true,urlcolor=blue,linkcolor=blue]{hyperref}
\pagestyle{empty}

\begin{document}

\noindent\textbf{Andrew H. Bond}, Senior Member, IEEE\\
Department of Computer Engineering, San Jos\'e State University\\
San Jos\'e, CA 95192, USA\\
\href{mailto:andrew.bond@sjsu.edu}{andrew.bond@sjsu.edu} \quad ORCID
\href{https://orcid.org/0009-0003-2599-6158}{0009-0003-2599-6158}

\medskip
\noindent July 2026

\bigskip
\noindent To the Editor-in-Chief\\
\emph{IEEE Transactions on Information Theory}

\bigskip
\noindent Dear Editor-in-Chief,

\medskip
I am pleased to submit the enclosed manuscript, ``Observation Theory: Geometry,
Distortion, and Reliability as Properties of Observation,'' for consideration as a
regular paper in the \emph{IEEE Transactions on Information Theory}.

The paper develops a consumer-relative theory of lossy representation. From a
second-order expansion of a downstream task's distortion it isolates a single
controlling quantity---the reconstruction error read through the consumer's read
operator $P_{\mathcal{C}}=J^\top G J$---and shows that ordinary mean-squared-error
reconstruction is its isotropic corner $P_{\mathcal{C}}=I$, the corner in which
Shannon's Gaussian rate--distortion function and Gauss least squares reside. The
principal contributions are information-theoretic. (i) A consumer-relative
rate--distortion function, with achievability and converse for a general source, and a
coding theorem for the \emph{true} nonlinear consumer divergence that reduces exactly to
output-side coding. (ii) For a Gaussian source read by two operators, the \emph{complete
two-stage rate region}: successive refinability is characterized by a single Loewner
condition---the observers' optimal error covariances must nest---with a closed-form
max-determinant rate loss when they do not, extended to the $k$-observer chain. (iii) Two
coding-theoretic consequences of the reduction: a finite-blocklength result whose effective
dimension is the \emph{rank} of the read operator rather than the ambient dimension, and a
source--channel separation theorem for the consumer distortion. (iv) An operational theory
of a \emph{misidentified} observer---mis-weighting a read direction costs a bounded rate
tax, whereas omitting one erects an irreducible distortion floor that no rate can cross---an
asymmetry that, within the Gaussian model, derives a compression catastrophe observed in
the experiments. (v) The identification of the read operator with the pullback geometry of
\emph{distinguishability}, and its recovery from black-box consumer queries, validated blind
against a planted ground truth. Throughout, ordinary reconstruction is the isotropic corner
$P_{\mathcal{C}}=I$, and the framework is positioned as the identifiable geometric middle
term between Shannon rate--distortion and the information bottleneck.

The work lies squarely within the Transactions' scope---Shannon theory,
rate--distortion and source coding, quantization, and information geometry, with
connections to learning and inference. It draws on and extends the indirect/remote
source-coding tradition (Dobrushin--Tsybakov, Wolf--Ziv, Witsenhausen), coding for
computing (K\"orner, Orlitsky--Roche), and the rate--distortion--perception and
information-bottleneck literatures, all of which are cited and discussed.

The material is original and previously unpublished; it has not been presented, in
whole or in part, at any conference, and it is not under review at any other journal.
It relates to, but is distinct from, two companion works by the author (on the
spectral/angular observer and on geometric key quantization) that are cited in the
manuscript; the present paper is a self-contained information-theoretic synthesis and
is not a verbatim extension of either.

In the interest of reproducible research, every empirical claim in the paper is
governed by a sealed, before-the-fact registration with a pre-declared acceptance bar.
The pre-registrations, experiment code, committed result artifacts, and a
continuous-integration workflow that re-runs the reproducible experiments and verifies
the paper's theoretical constructions are publicly available at the repository named in
the manuscript, with a permanent archival snapshot deposited at Zenodo
(\href{https://doi.org/10.5281/zenodo.21423316}{doi:10.5281/zenodo.21423316}). In addition,
each analytic result was independently re-derived in a fresh-context verification pass whose
outcome---including two self-corrections caught before submission---is recorded in the
ledger. Consistent with the Society's encouragement, I intend to post a preprint to arXiv at
the time of submission.

I confirm that I am the sole author; that the submission is prepared in the required
single-column format and is well within the 50-page review limit; and that I hold a
registered ORCID. I would be glad to suggest qualified reviewers or to provide any
additional material the editors may find helpful.

Thank you for your time and for considering this work.

\bigskip
\noindent Sincerely,

\medskip
\noindent Andrew H. Bond, Senior Member, IEEE\\
Department of Computer Engineering, San Jos\'e State University\\
\href{mailto:andrew.bond@sjsu.edu}{andrew.bond@sjsu.edu}

\end{document}
