\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\newcommand{\PC}{P_{\mathcal{C}}}
\newcommand{\Sd}{\Sigma_{\delta}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\dO}{d_{\mathcal{O}}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}

\title{\textbf{Observation Theory}\\[2pt]
\large Geometry, Distortion, and Reliability as Properties of Observation}
\author{Andrew H. Bond\\ \small \texttt{agi.hpc@gmail.com}}
\date{July 2026 --- \emph{Preprint (registration-first; every empirical claim resolves to a sealed ledger row).}}

\begin{document}
\maketitle

\begin{abstract}
We unify and operationalize a family of results under one thesis---\emph{Observation
Theory}: \emph{operational} geometry, distortion, and reliability are not properties
of the object alone but of the \emph{relation} among object, observer (a consumer and
its output metric), channel, and budget. From a second-order expansion of the downstream metric we recover
the leading-order controlling quantity---the reconstruction error read through the
consumer, $\dO=\E_x\tr(\PC(x)\Sd(x))$, reducing under stated conditions to
$\tr(\PC\Sd)$---with $\Sd$ the error second moment and $\PC=\E[J^\top G J]$ the
output-metric-weighted sensitivity. At root this is a \emph{distinguishability}:
$\PC=J^\top G J$ is the pullback of the consumer's output metric $G$ (the Fisher
metric for KL and kindred statistical divergences), so each observer
induces a geometry and a quotient $X/\!\sim_{\mathcal{C}}$ of the representation by
what it cannot tell apart. This weighted-quadratic fidelity is classical:
it is the indirect/remote source-coding distortion of Dobrushin--Tsybakov and
Wolf--Ziv, reduces to Shannon's Gaussian rate--distortion by Berger's
change of variables, and is Hessian-weighted loss-aware compression (OBD/OBS through
GPTQ) written as a metric. Our contribution is the \emph{middle term}: $\PC$ is a
geometric, PSD, water-fillable operator that is \emph{recoverable from black-box
consumer queries under a specified observation distribution and output metric},
sitting between Shannon's Gaussian Euclidean-MSE specialization ($\PC=I$) and the
information bottleneck. We make that middle term testable and test it, under sealed, before-run
registrations with ex-ante bars: distortion---$\tr(\PC\Sd)$ governs downstream error,
demonstrated and replicated across softmax attention, retrieval ranking, and
gradient optimization, with a reconstruction-identical code \emph{flipping} its
verdict when the consumer changes; identifiability---blind planted-subspace recovery
at $16\times$ chance, validated against ground truth; capacity---an extreme-value
vacuity threshold that locates retrieval death to a few percent across $14$ corpora;
and rate---a fixed-budget verdict that \emph{inverts} under budget-matched
observation on real embedding manifolds. The analytic consumers are \emph{synthetic}
by design (planted read operators, chosen query banks, constructed Hessians),
including an $H{=}I$ control where reconstruction becomes exactly optimal; the one
real-substrate face is the budget inversion. A prospective, preregistered attempt to
harden on a \emph{trained} consumer (a Llama-3.2-3B attention layer, GO-P-2026-020)
\emph{misses} its sealed bar: the blind probe recovers the real read operator at
$4.5\times$ chance but below the $0.60$ overlap threshold ($0.567$), and---because the
registered quantizer pair was not reconstruction-matched---the consumer-projected
predictor does not \emph{beat} reconstruction, so the decisive dissociation is
untested on the trained model. On this evidence the theory is bounded to synthetic
consumers; a recon-matched, point-dependent rematch is the open decisive validation. A fifth conjecture---a density quotient restoring
the invariant in non-spectral retrieval---is \emph{refuted} across four prospective
tests and carried as a boundary. For the fixed (or averaged) weighted-quadratic
model we prove achievability and converse of the consumer-relative rate--distortion
function for a general source (Appendix~\ref{app:coding}), and for discrete sources
exhibit a consumer-lossless floor below the source entropy. We do \emph{not} claim
Shannon was incomplete---his framework already admits any fidelity criterion, and
$R_{\mathcal{C}}(D)$ is a classical rate--distortion problem once the consumer's
distortion is chosen. Euclidean reconstruction is its \emph{isotropic-observer}
corner ($\PC=I$); Observation Theory's contribution is to make that distortion
measure an \emph{identifiable geometric object}, recover it from the consumer, and
show adversarially that changing it changes which representation is optimal.
\end{abstract}

\section{Introduction}

Shannon fixed a distortion measure and asked for the fewest bits meeting it
\cite{shannon1948,shannon1959}; Gauss fixed the same measure and asked for the
least-squares estimate \cite{gauss1809}. This paper reads their common assumption
as the variable of interest: the error that matters is not the one a code makes but
the one a consumer \emph{reads}, and different consumers read different subspaces of
the same error. Once the distortion is the task's, Euclidean reconstruction is one
corner---the isotropic observer $\PC=I$---and any anisotropic or rank-deficient $\PC$
reads the error differently.

The idea that one should code a source to serve a \emph{functional} of it, not the
source, is not new: it is the indirect (remote) rate--distortion problem
\cite{dobrushin1962,wolfziv1970,witsenhausen1980}; its lossless, discrete form is
coding for computing and graph entropy \cite{korner1973,orlitsky2001}; its modern
face is the rate--distortion--perception tradeoff \cite{blau2019} and task-oriented
/ semantic communication \cite{gunduz2023}; and its deployed form, for neural
networks, is loss-aware (Hessian-weighted) compression, from Optimal Brain
Damage/Surgeon \cite{lecun1990,hassibi1993} to Hessian-aware quantization and GPTQ
\cite{dong2019hawq,frantar2023gptq}. What is missing across these is a single,
\emph{identifiable geometric} object that names the task's metric and lets one
predict, recover, and test it. Observation Theory supplies it: the read operator
$\PC$, the consumer's local output metric, sitting as the geometric middle term
between Shannon's Gaussian Euclidean-MSE specialization ($\PC=I$) and the information
bottleneck \cite{tishby1999} (an arbitrary, non-geometric relevance variable). We
contribute (i) the map placing these literatures on one axis; (ii) $\PC$'s
\emph{identifiability}---recovery from black-box consumer queries under a specified
observation distribution, validated blind against a
planted ground truth, which the quantization literature (estimating the Hessian from
calibration data) never validates against truth; (iii) an ex-ante extreme-value
reliability certificate; (iv) budget-relativity of the verdict; and (v) a
registration-first empirical program---sealed predictions, ex-ante bars, a refuted
face carried as prominently as the wins---whose recurring demonstration is a single
code whose downstream verdict \emph{inverts} when nothing changes but the consumer.
The three companion works are Paper~I (the angular observer and the spectral
budget-inversion mechanism) \cite{paper1}, Paper~II (geometric key quantization and
the reconstruction-is-not-the-target negative) \cite{paper2}, and the claim ledger
\cite{ledger2026}; this is their information-theoretic synthesis.

\section{The controlling quantity}\label{sec:framework}

\begin{definition}[Instance]
An instance is $(X,\mathcal{C},\mathcal{Q})$: a representation with vectors
$x\in\R^{d}$, a \emph{consumer} $\mathcal{C}:\R^{d}\to\R^m$ with a divergence $D_Y$
on its outputs, and a family $\mathcal{Q}$ of lossy codes producing $\hat x=x+\delta$.
The downstream distortion is the pointwise random quantity
$\dO:=D_Y(\mathcal{C}(x),\mathcal{C}(\hat x))$, measured on the consumer output, never
assumed from the ambient metric of $X$; write $\bar{\dO}:=\E[\dO]$ for its expectation
over the instance.
\end{definition}

\begin{proposition}[Consumer-projected distortion]\label{prop:main}
Let $D_Y$ be twice differentiable with $D_Y(u,u)=0$, $\nabla_v D_Y(u,v)|_{v=u}=0$,
and \emph{local output metric} $G(u)=\nabla^2_v D_Y(u,v)|_{v=u}\succeq0$ (the Fisher
information metric when $D_Y$ is the KL divergence or a kindred statistical
divergence, a generic local metric otherwise); let
$J(x)=\partial\mathcal{C}/\partial x$. For error $\delta=\hat x-x$ with conditional
\emph{second moment} $\Sd(x)=\E[\delta\delta^\top\mid x]$, the leading-order expected
distortion is
\[
\bar{\dO} \;=\; \tfrac12\,\E_x\tr\!\big(\PC(x)\,\Sd(x)\big)\;+\;r,
\qquad \PC(x)=J(x)^\top G(\mathcal{C}(x))\,J(x)\succeq0 ,
\]
so $\PC$ is the pullback of the local output metric $G$ along $\mathcal{C}$. Making
the asymptotics explicit: for a small-noise family $\delta_\varepsilon=\varepsilon Z$
with $\E\|Z\|^2<\infty$ and the second-order Taylor remainder dominated uniformly on a
neighborhood of the diagonal, $\bar{\dO}=\tfrac{\varepsilon^2}{2}\,\E[Z^\top\PC(X)Z]
+o(\varepsilon^2)$ (i.e.\ $r=o(\varepsilon^2)$).
When $\Sd(x)$ is uncorrelated with the point-dependent weight $\PC(x)$ this factors
as $\tfrac12\tr(\PC\Sd)$ with $\PC=\E_x[\PC(x)]$, $\Sd=\E_x[\Sd(x)]$. The controlling
object is the \emph{second} moment, not the covariance: pointwise
$\tr(\PC(x)\Sd(x))=\tr(\PC(x)\,\mathrm{Cov}[\delta\mid x])+\mu_\delta(x)^\top\PC(x)
\mu_\delta(x)$ with $\mu_\delta(x)=\E[\delta\mid x]$, then averaged over $x$, so a
biased code's offset is \emph{part} of the distortion, not averaged away---which is
why quantizer bias is first-class (\S\ref{sec:readop}). Reconstruction distortion
$\E\|\delta\|^2=\tr(\Sd)$ is the \emph{isotropic} corner $\PC=I$.
\end{proposition}
\begin{proof}
Expand $\mathcal{C}(x+\delta)=\mathcal{C}(x)+J\delta+O(\|\delta\|^2)$ and $D_Y$ to
second order about its minimum; the value and gradient terms vanish, leaving
$\tfrac12\delta^\top(J^\top GJ)\delta+o(\|\delta\|^2)$. Take $\E_x$; the factored
form requires the stated uncorrelatedness (for real quantizers both $J$ and $\Sd$
depend on $x$, so the exact object is the $x$-averaged trace).
\end{proof}

We keep $\E_x\tr(\PC(x)\Sd(x))$ as primary and use the factored $\tr(\PC\Sd)$ as the
convenience it is. (We drop the harmless $\tfrac12$.)

\subsection*{Distinguishability, and the observer's quotient}

The form $\delta^\top\PC(x)\delta=(J\delta)^\top G\,(J\delta)$ is not, at root, a
compression cost; it is a \emph{distinguishability}. It measures how far the
consumer's outputs $\mathcal{C}(x)$ and $\mathcal{C}(x{+}\delta)$ separate in the
output metric $G$, and when $G$ is the Fisher information metric of the output
distribution, $\delta^\top\PC\delta$ \emph{is} the local statistical
distinguishability (the leading KL / $\chi^2$) of those outputs. So $\PC$ is the
pullback through $\mathcal{C}$ of the geometry of distinguishability: Observation
Theory is pullback geometry.

Two consequences follow, and both recur below. First, define the global operational
equivalence $x\sim_{\mathcal{C}}x'\iff D_Y(\mathcal{C}(x),\mathcal{C}(x'))=0$
(statistical indistinguishability of the outputs). Its infinitesimal shadow is
$\ker\PC(x)=\ker\!\big(G^{1/2}J\big)$, which equals $\ker J$ when $G\succ0$ on
$\operatorname{im}J$ but is larger when $G$ is only semidefinite---the output can move
parametrically yet stay indistinguishable. Under a constant-rank / integrability
condition on $\PC(x)$ this null distribution integrates to leaves tangent to
$\ker\PC(x)$; that the \emph{global} divergence vanishes along each leaf---so the
leaves are exactly the fibers of $\sim_{\mathcal{C}}$---needs a further compatibility
condition, namely $D_Y(\mathcal{C}(x),\mathcal{C}(x'))=0$ along them (equivalently,
$\mathcal{C}$ factors through $X/\!\sim_{\mathcal{C}}$). Where it holds, the
consumer's true object is the quotient $X/\!\sim_{\mathcal{C}}$, not $X$:
perturbations in $\ker\PC$ exist in the representation but have \emph{no operational
consequence}. In the fixed-$\PC$ model of \S\ref{sec:rd} this quotient is represented
by $\Pi X$ and Prop.~\ref{prop:floor} gives the zero-distortion rate
$R_{\mathcal{C}}(0)=H(\Pi X)$ (the point-dependent case is open)---one pays only for
what one can tell apart. Second, $\ker\PC$ is
gauge-like \emph{but only relative to $\mathcal{C}$}: another consumer may read those
directions, so it is not a true gauge redundancy (which would be invisible to every
admissible observer). One observer's noise is another's signal; one observer's state
variable is another's gauge freedom. Adjoining the budget of \S\ref{sec:rate}, an
observer is a triple $\mathcal{O}=(\mathcal{C},G,B)$---readout, output metric, and
resolution---and the geometry it resolves is relative to all three.

\begin{quote}
\textbf{Principle of observational relativity.} Operational fidelity, distortion,
and resolved geometry are not properties of the object alone; they are properties of
the relation among object, observer $(\mathcal{C},G)$, channel, and budget, and are
ill-posed until all are specified. Observation induces geometry; each observer pulls
back a world.
\end{quote}

The rest of the paper derives $\PC$ for concrete consumers, shows it recoverable
from queries, and studies what the channel (\S\ref{sec:capacity}) and budget
(\S\ref{sec:rate}) do to the geometry it induces.

\section{The read operator, derived}\label{sec:readop}

\begin{proposition}[Softmax attention]\label{prop:softmax}
Read keys $\{k_s\}_{s=1}^{S}$ through a query bank: logits $l_s=q^\top k_s/\sqrt{d}$,
output $p=\mathrm{softmax}(l)$, $D_Y=\mathrm{KL}$. Under key perturbation $\delta_s$,
\[
\bar{\dO}=\tfrac{1}{2d}\,\E_q\,\mathrm{Var}^{\,p}_s\!\big(q^\top\delta_s\big)
+O(\|\delta\|^3)
=\tfrac{1}{2d}\,\E_q\tr\!\big(qq^\top A_q\big)+O(\|\delta\|^3),
\qquad A_q=\sum_s p_s(q)\,\tilde\delta_s\tilde\delta_s^\top,
\]
where $\tilde\delta_s=\delta_s-\sum_{s'}p_{s'}\delta_{s'}$ is the across-token
$p$-demeaned error and $A_q$ its $p(q)$-weighted second moment. This
\emph{factors} to $\tfrac{1}{2d}\tr(Q\,\bar\Sigma_\delta)$, with $Q=\E_q[qq^\top]$
and $\bar\Sigma_\delta=\E_q[A_q]$, \emph{only} when $A_q$ is uncorrelated with
$qq^\top$ (the proviso of Prop.~\ref{prop:main}); in general the expectation stays
coupled, since $p$ depends on $q$.
\end{proposition}
\begin{proof}
The Fisher metric of $\mathrm{softmax}$ is $G=\mathrm{diag}(p)-pp^\top$, so
$\mathrm{KL}(p\|\mathrm{softmax}(l+\Delta l))=\tfrac12\mathrm{Var}^p_s(\Delta l_s)
+O(\Delta l^3)$ with $\Delta l_s=q^\top\delta_s/\sqrt d$. The $-pp^\top$ term is the
softmax's invariance to a common logit shift; it removes the $p$-mean, giving
$\tilde\delta_s$. Since $p$ depends on $q$, the outer $\bar\Sigma_\delta$ carries the
$\E_q$ and the final $Q$-factorization inherits the uncorrelatedness proviso of
Prop.~\ref{prop:main}.
\end{proof}
Two consequences are the phenomena we later measure. The feature-space factor is
generated by $qq^\top$, reducing to $Q=\E_q[qq^\top]$ under the decorrelation /
uniform-weight regime above; so an isotropic bank yields an isotropic feature
operator (reconstruction-like) while a subspace bank projects---the read operator,
hence the winning code, changes with the query distribution. And only the
\emph{demeaned} error survives, so the across-token mean is a nuisance the softmax
cannot see. This last fact resolves an apparent tension with the DC-offset
collapse of Paper~II \cite{paper2}: a truly common error component is free by shift
invariance, but a per-channel DC \emph{offset} is not a common component---it inflates
the quantizer's range, wasting levels and enlarging the \emph{visible} per-token
error, which is exactly why an asymmetric zero-point (asym-NF4) repairs it at no bit
cost. Invisible-to-the-consumer and free-to-encode are the same statement; range
waste is a different one.

\begin{proposition}[Gradient-descent optimizer; exact]\label{prop:opt}
Let $L(\theta)=\tfrac12\theta^\top H\theta$ with $H\succeq0$, $g=H\theta$, and
$\hat g=g+\delta$ with conditionally unbiased error $\E[\delta\mid\theta]=0$, used in
a step of size $\eta$. Then $L(\theta-\eta\hat g)-L(\theta-\eta g)=\tfrac12\eta^2\,
\delta^\top H\delta-\eta\,\delta^\top H(\theta-\eta g)$, whose error-quadratic part
is $\tfrac12\eta^2\tr(H\,\delta\delta^\top)$; the linear cross term vanishes in
expectation ($\E[\delta\mid\theta]=0$), giving $\E[\dO]=\tfrac12\eta^2\tr(H\,\Sd)$,
so $\PC=H\succeq0$ \emph{exactly}.
\end{proposition}
\noindent This is loss-aware pruning/quantization made a metric: Hessian-weighted
deletion \cite{lecun1990,hassibi1993,dong2019hawq,frantar2023gptq} \emph{is}
$\tr(H\Sd)$, and calibration-based Hessian estimation is the production form of the
probe of \S\ref{sec:id}. Reconstruction is the flat case $H=I$. For a non-convex
loss the Hessian may be indefinite; the metric reading then takes $\PC$ to be the
PSD Gauss--Newton / Fisher curvature in place of $H$.

\begin{remark}[Retrieval]
For score-ranking $s(q,x_i)=q^\top x_i$, the inverted-pair fraction is, to leading
order for smooth rank statistics, monotone in
$\tr(\PC\bar\Sigma_\delta)/\tr(\PC\bar\Sigma_0)$, with $\PC=\E[qq^\top]$,
$\bar\Sigma_\delta=\mathrm{Cov}_i(\delta_i)$ the across-item error covariance and
$\bar\Sigma_0=\mathrm{Cov}_i(x_i)$ the signal covariance (both demeaned across items);
$\PC$ is a subspace projector when queries concentrate. Rank statistics are
non-smooth; we treat this instance empirically (\S\ref{sec:distortion}).
\end{remark}

\section{Distortion: the consumer is the metric (GO-2)}\label{sec:distortion}

Proposition~\ref{prop:softmax} predicts that at matched reconstruction the ranking of
two codes is set by $\tr(\PC\Sd)$, hence by the query distribution.

\paragraph{Dissociation.} Two 4-bit codes at an audited matched budget (spread
$0.19$ bit) with \emph{identical} reconstruction ($0.0938$ vs $0.0934$): the
whole-vector code is $2.53\times$ worse on softmax-KL ($12/12$ seeds); a
wrong-quotient control is $21\times$ worse at $\sim\!5\times$ the reconstruction.
Reconstruction is not the controlling quantity (\texttt{[demonstrated]}).

\paragraph{Flip.} Holding reconstruction fixed and changing only the query bank, the
ranking \emph{inverts}: isotropic $\Rightarrow$ code $a$ better ($2.5\times$,
$12/12$); subspace $\Rightarrow$ code $b$ better ($12/12$). The directly computed
$\tr(\PC\Sd)$ tracks the sign on every seed in both consumers; reconstruction, a
consumer-blind scalar, cannot (rank correlation $0.40$ vs $1.00$). This is
Proposition~\ref{prop:softmax} with $Q$ swapped (\texttt{[demonstrated]}).

\paragraph{Replication and out-of-sample.} The flip replicates on retrieval ranking
(different representation, set-valued consumer): identical reconstruction
($0.0964$), ranking flips with the read subspace (log-advantage $-4.70$ vs $+4.65$,
$12/12$), $\tr(\PC\Sd)$ tracking, reconstruction blind (\texttt{[replicated]}). It
generalizes to optimization with $\PC=H$ (Prop.~\ref{prop:opt}): at matched
reconstruction the curvature-metric distortion flips between two Hessians ($12/12$),
tracked by $\tr(H\Sd)$, reconstruction uninformative ($-0.20$)---and under $H=I$
reconstruction tracks \emph{perfectly} ($1.00$), the derived $\PC=I$ corner observed
(\texttt{[predicted]}, out-of-sample). One derivation; three read operators. These
consumers are \emph{synthetic}---the point is to validate the theory's logic (the
$H=I$ positive control being the clean confirmation); the trained-model hardening
(\S\ref{sec:discussion}), attempted prospectively, does \emph{not} yet clear its
sealed bar and remains the outstanding validation.

\section{Identifiability: the read subspace is recoverable (GO-1)}\label{sec:id}

\begin{proposition}[Recovery of the read subspace]\label{prop:id}
Let $\mathcal{C}(x)=f(Wx)$, $W\in\R^{r\times d}$ of rank $r$, $f$ coordinatewise with
$\E[f_i'(Wx)^2]>0$ for every coordinate $i$ (no saturated unit). Then
$\E_x[J^\top J]=W^\top\E[\mathrm{diag}(f'^2)]W$ has rank $r$ with column space
$\operatorname{row}(W)$, so its top-$r$ eigenvectors span the read subspace; a
finite-difference estimate $\hat J u\approx(\mathcal{C}(x{+}hu)-\mathcal{C}(x{-}hu))/2h$
recovers it from consumer queries alone.
\end{proposition}

We built such a probe---receiving only the callable $\mathcal{C}(\cdot)$ and the
dimensions, never $W$, the codes, or $\dO$ (blinding enforced in code). It recovers
the planted read subspace at overlap $0.936$ against a $0.059$ chance baseline
($\approx16\times$), and the \emph{blind} $\tr(\hat\PC\Sd)$ predicts the flip of
\S\ref{sec:distortion} on $12/12$ seeds in both consumers with rank correlation
$1.00$---identical to ground truth---while reconstruction cannot ($0.40$)
(\texttt{[predicted]}). Two scope notes. Prop.~\ref{prop:id} recovers the read \emph{subspace} for the
$f(Wx)$ model with a Euclidean output metric; recovering the \emph{full} operator
$\E_x[J^\top G J]$---its eigenvalues, and a general $G$---is the broader program.
And $\PC$ depends on the probe-point distribution and on $G$, so it is identifiable
from black-box consumer queries \emph{under a specified observation distribution},
not from the callable alone. Loss-aware compression estimates such a metric from
calibration data; the contribution here is validating \emph{recovery against a
planted ground truth under enforced blinding}, and composing it: from consumer
queries under that distribution one predicts which code wins, before compressing
anything.

\section{Capacity: a vacuity threshold, derived (GO-3)}\label{sec:capacity}

\begin{proposition}[Retrieval vacuity threshold]\label{prop:capacity}
Model the $N\!-\!1$ distractor scores, standardized to mean $0$, variance $1$, as
i.i.d.\ standard Gaussian (a reference model). Since $\Pr[M_{N-1}\le m]=\Phi(m)^{N-1}$,
top-1 success $\Pr[\hat\mu>M_{N-1}]$ crosses $\tfrac12$ at the \emph{exact median}
of the maximum,
\[
\mu_{\mathrm{crit}}:=\operatorname{med}(M_{N-1})=\Phi^{-1}\!\big(2^{-1/(N-1)}\big),
\qquad \text{leading asymptotic }\sqrt{2\ln(N{-}1)},
\]
where $\hat\mu=(\overline{\text{true}}-\overline{\text{distr}})/\mathrm{std}$ is the
margin certificate and $\rho=\hat\mu/\mu_{\mathrm{crit}}$; the certificate is
\emph{vacuous} for $\rho<1$.
\end{proposition}
\begin{proof}[Derivation]
$\Pr[M_{N-1}\le m]=\Phi(m)^{N-1}$; setting it to $\tfrac12$ gives $m=\Phi^{-1}(2^{-1/(N-1)})$,
the exact median. The Gumbel centering is $\alpha_N=\sqrt{2\ln(N{-}1)}-b_N(\ln\ln(N{-}1)
+\ln4\pi)/2$ with scale $b_N=1/\sqrt{2\ln(N{-}1)}$; the crude $\sqrt{2\ln(N{-}1)}$
omits the $-b_N(\ln\ln(N{-}1)+\ln4\pi)/2$ correction and so \emph{overshoots}. Median
and mean share this centering and differ at order $b_N$:
$\operatorname{med}(M_{N-1})=\alpha_N-b_N\ln\ln2+o(b_N)$ and
$\E[M_{N-1}]=\alpha_N+\gamma b_N+o(b_N)$ ($\gamma$ Euler--Mascheroni), so the mean
exceeds the median by $(\gamma+\ln\ln2)\,b_N$. The recall-$\tfrac12$ crossing is the
exact median $\Phi^{-1}(2^{-1/(N-1)})$---not the mean, not the leading order.
\end{proof}

Across $14$ structurally distinct corpora the certificate orders them by recall at
rank correlation $0.991$, beating the un-normalized margin ($0.873$), with a
finite-width transition (\texttt{[demonstrated]}). The registered harness used the
\emph{mean} $\E[M_{N-1}]$ (Monte-Carlo) as its $\mu_{\mathrm{crit}}$ reference, and
the recall-$\tfrac12$ crossing sits at $\rho=0.948$---slightly \emph{below} $1$
against that reference. This is expected and quantitative: the exact $\tfrac12$
threshold is the \emph{median}, below the mean by $\Theta(1/a_N)$, so the $\sim\!5\%$
shortfall is largely the mean--median offset (against the exact median of this
proposition the crossing sits at $\approx1$). Positive dependence among distractors
\emph{can} lower it further by reducing the effective count; heavier tails, by
contrast, \emph{raise} the maximum---the two must not be grouped. The exact
$\tfrac12$ crossing further assumes a \emph{fixed} standardized true margin and
i.i.d.\ Gaussian distractors; when the true score is itself random or correlated with
the distractors, $\rho$ is a reference certificate rather than an exact recall
threshold. The \emph{form} of the threshold is nonetheless robust: the
$\sqrt{2\ln N}$ scaling is the centering of the Gumbel max-domain of attraction,
shared by every distractor law with sub-exponential (lighter-than-polynomial)
tails---Gaussian, exponential, Weibull-type---so non-Gaussianity \emph{within} that
domain moves the constants $(\alpha_N,b_N)$, not the scaling law; positive correlation
enters as an \emph{effective count}, replacing $N$ by $\theta N$ through the extremal
index $\theta\in(0,1]$ and shifting $\mu_{\mathrm{crit}}$ by
$\tfrac{\ln\theta}{2\sqrt{2\ln N}}$ at leading order; and only genuinely heavy
(Fr\'echet-domain, polynomial) tails change the scaling itself---the certificate's
stated failure mode. The $14$-corpus ordering above ($\rho=0.99$ on clustered,
non-Gaussian real embeddings) is the empirical counterpart: real structure shifts the
absolute death point but preserves the ordering the certificate is used for.
Capacity says, from a computable quantity, \emph{where} the channel dies
before any retrieval is run.

\section{Rate: the verdict belongs to the budget (GO-4)}\label{sec:rate}

The read operator, capacity, and distortion are stated at a fixed observer budget;
the budget is a rate, and it inverts the verdict. The mechanism is spectral: an
eigenfunction $\varphi_k$ with eigenvalue $\lambda_k$ oscillates on the wavelength
$\lambda_k^{-1/2}$, and Weyl's law $N(\lambda)\sim c\,\mathrm{Vol}(M)\lambda^{d_M/2}$
($d_M$ the intrinsic dimension) grows the count of modes below a fixed scale with
resolution.

\begin{lemma}[Fixed-band consistency]\label{lem:band}
At a fixed low-mode budget $m$, as the sample grows the graph operator converges
spectrally to the manifold operator \cite{calder2022}, so the fixed low-$\lambda$
band sharpens and geodesic fidelity \emph{converges toward its continuum value} and is
predicted to improve asymptotically (the spectral-convergence mechanism of
Paper~I \cite{paper1}). Convergence does not force step-wise monotonicity in $N$; the
monotone finite-sample trend below is the \emph{registered empirical} result, not a
consequence of the lemma.
\end{lemma}

Holding instead the resolved scale $\tau$ fixed (spectral-gap matching), $m=N(\tau)$
grows with $N$; the newly resolved modes near $\tau$ have wavelength approaching the
sampling scale and, for a non-ideal substrate, inject sub-geodesic detail, so
geodesic fidelity is nonincreasing---Paper~I's registered inversion \cite{paper1}. We
replicate it on a new substrate: $768$-dimensional multilingual book-paragraph
embeddings (Project Gutenberg translations; $32{,}000$ points, three seeds). Fixed
$m=10$: geometry preservation rises with $N$ (trend $+0.4,+1.0,+1.0$).
Spectral-gap-matched, $m:10\to121/126/159$: geometry \emph{collapses} (trend $-1.0$
every seed); the budgets agree at the smallest $N$ (gap $\le0.002$)
(\texttt{[replicated]}). Opposite verdicts about the same data, set by the observer's
rate: ``clean geometry'' is a relation between observer resolution and substrate
scale, not a substrate invariant.

\section{A consumer-relative rate--distortion function}\label{sec:rd}

The consumer-projected distortion is the single-letter measure
$d_{\mathcal{C}}(x,\hat x)=(x-\hat x)^\top\PC(x-\hat x)$ for a \emph{fixed} $\PC$
(constant, or the average $\E_x[\PC(x)]$ under the proviso of Prop.~\ref{prop:main});
define $R_{\mathcal{C}}(D)=\inf\{I(X;\hat X):\E\,d_{\mathcal{C}}(X,\hat X)\le D\}$.
This fixed-$\PC$ model must be kept distinct from the \emph{local} observation metric
$\PC(x)$ of Prop.~\ref{prop:main}: for a nonlinear consumer $\PC(x)$ varies, there is
no single tilted source---a direction invisible at one point can be visible at
another---and the coding theorem for that case is open. The results below, and
Appendix~\ref{app:coding}, are for the fixed weighted-quadratic model, whose theory
is classical \cite{berger1971}.

\begin{proposition}[Gaussian rate--distortion; Berger reduction]\label{prop:rd}
For $X\sim\mathcal{N}(0,\Sigma_x)$, in the basis diagonalizing
$\PC^{1/2}\Sigma_x\PC^{1/2}$ with eigenvalues $\gamma_i$,
\[
R_{\mathcal{C}}(D)=\sum_i \tfrac12\log_2^{+}\!\frac{\gamma_i}{\theta},\qquad
\sum_i\min(\gamma_i,\theta)=D,
\]
a reverse water-filling on the $\PC$-weighted source spectrum. For $\PC=I$ this is
Shannon's Gaussian $R(D)$ \cite{shannon1959}; the estimation dual is Gauss--Markov
least squares \cite{gauss1809}.
\end{proposition}
\begin{proof}
$\tr(\PC\Sd)=\|\PC^{1/2}(x-\hat x)\|^2$; with $y=\PC^{1/2}x$ this is MSE on
$y\sim\mathcal{N}(0,\PC^{1/2}\Sigma_x\PC^{1/2})$, and Shannon water-filling applies.
The transform is invertible off $\ker\PC$, on which the consumer is blind and no rate
is spent.
\end{proof}

Bits water-fill on directions the consumer reads \emph{and} the source varies,
spending nothing on $\ker\PC$---why a reconstruction-optimal code ($\PC=I$
allocation) is beaten by a consumer-matched one when $\PC\ne I$, and why the winner
\emph{flips} with $\PC$ (\S\ref{sec:distortion}). The $H=I$ control confirms it: only
at the isotropic $\PC=I$ is reconstruction the right objective, and there it is
exactly right.

\begin{proposition}[Consumer-lossless floor; discrete source]\label{prop:floor}
Let $X$ be discrete with finite entropy, $\Pi$ the projector onto
$\operatorname{range}(\PC)$. There exist codes with $\tr(\PC\Sd)=0$ at rate
\[
R_{\mathcal{C}}(0)=H(\Pi X)\le H(X),\qquad H(X)-R_{\mathcal{C}}(0)=H(X\mid\Pi X)\ge0,
\]
with equality iff $\Pi$ is injective on $\operatorname{supp}(X)$; the saving is
strict exactly when distinct symbols differ only within the consumer-invisible
subspace $\ker\PC$.
\end{proposition}
\begin{proof}
\emph{Achievability:} $\tr(\PC\Sd)$ depends on $\delta$ only through $\Pi\delta$;
sending $\Pi X$ losslessly and reconstructing $(\mathrm I-\Pi)X$ by its conditional
mode gives $\Pi\delta=0$, zero distortion, at rate $H(\Pi X)$. \emph{Converse:} zero
distortion forces $\Pi\hat X=\Pi X$ a.s., so $I(X;\hat X)\ge I(\Pi X;\Pi\hat X)=
H(\Pi X)$. Hence $R_{\mathcal{C}}(0)=H(\Pi X)$; if $\Pi$ collapses two support points,
$H(\Pi X)<H(X)$.
\end{proof}
\begin{remark}[Continuous sources]
When the tilted source $Y=\PC^{1/2}X$ has a nondegenerate continuous component,
lossless-to-the-consumer coding costs infinite rate ($R_{\mathcal{C}}(D)\to\infty$ as
$D\to0$, consistent with Prop.~\ref{prop:rd}); if $X$ lies entirely in $\ker\PC$ then
$Y$ is degenerate and the floor is zero. The surviving statement is
that at \emph{every} $D>0$ the water-filling of Prop.~\ref{prop:rd}, equivalently the
reduction of Cor.~\ref{cor:coding}, spends zero rate on $\ker\PC$: the operational
rate depends only on the $r$-dimensional tilted source. This lossless-to-the-consumer
phenomenon is exactly coding for computing and graph entropy in the discrete world
\cite{korner1973,orlitsky2001}, here given a geometric read operator.
\end{remark}

These form a \emph{conceptual axis}, not a formal nesting: Shannon rate--distortion
($\PC=I$, the isotropic observer), consumer-relative rate--distortion ($\PC$ a fixed
geometric, \emph{identifiable} operator, Props.~\ref{prop:rd},\ref{prop:floor}), and
the information bottleneck \cite{tishby1999}---which optimizes a different object (a
stochastic relevance variable) and to which the consumer-relative term connects only
through the local-Gaussian relevance limit ($Y=\mathcal{C}(X)$ deterministic,
second-order). We claim no containment theorem. The middle term is the operational
one---$\PC$ is PSD, water-fillable, and recoverable from the consumer
(Prop.~\ref{prop:id})---and its coding theorem (achievability and converse, fixed
$\PC$, general source) is Appendix~\ref{app:coding}.

\section{A boundary: the refuted quotient (GO-5)}\label{sec:boundary}

Not every plausible face survives a sealed prediction. In the spectral setting the
$\alpha{=}1$ anisotropic normalization of a diffusion operator cancels sampling
density and recovers density-free geometry \cite{coifman2006}: there the density mode
lies in $\ker\PC$. We conjectured this density quotient likewise restores the
invariant in a \emph{non-spectral} domain (locally-scaled ADC retrieval, a hubness
correction \cite{radovanovic2010}). It does not, across four prospective tests
(direct affinity twice, ADC on real embeddings, non-spectral diffusion-distance).
The reason is Proposition~\ref{prop:rd}: for a direct inner-product consumer the
density is \emph{read}---it is not in $\ker\PC$---so quotienting it removes signal and
manufactures anti-hubs; in the diffusion domain the direction is right ($\alpha{=}1$
optimal, recovered non-spectrally) but the effect is under $2\%$ and \emph{not
density-specific} (a uniform-density control gains as much). Per its sealed
registration the conjecture is \texttt{[refuted]}. The sharpened statement---the
$\alpha{=}1$ cancellation is a property of the diffusion operator and does not
transfer to direct observation---is Proposition~\ref{prop:rd} read backwards: rate is
saved on $\ker\PC$ only when the nuisance actually lies there. The refutation
sharpens the theory rather than embarrassing it.

\section{Honest negatives and method}\label{sec:negatives}

Eleven standing negatives bound the theory and are reported with the prominence of
the positives. Reconstruction cosine is not a proxy for key quality (Paper~II
\cite{paper2}, the motivating negative). The flip is observable only for
reconstruction-\emph{matched} codes: absent a tie one code Pareto-dominates and
reconstruction is not falsified (a prospective miss locating a precondition,
consistent with the water-filling of Prop.~\ref{prop:rd}). The projected-variance
trace governs the discriminating comparison but is not a complete rank statistic over
all codes. The density quotient (\S\ref{sec:boundary}). Two are methodological
confessions: twice a substantive claim passed at $12/12$ while an over-strict
secondary gate we had written failed, and twice a degenerate substrate was caught
only by a regime-coverage gate.

Every empirical claim was governed by a dated, hashed registry entry
committed---sealed---before the first measurement it governs; bars were fixed ex
ante; misses enter the ledger as misses; amendments only dated and logged before
unblinding; prospective claims (\S\ref{sec:id},\ref{sec:rate},\ref{sec:boundary}) were
blinded where feasible. The house rule is that the paper may not assert what the
ledger cannot show.

\section{Discussion}\label{sec:discussion}

In one line: Observation Theory identifies operational geometry with the pullback
geometry of distinguishability, then makes that geometry recoverable and
experimentally consequential. It asks a change of default. Where a pipeline minimizes
reconstruction and hopes the task follows, the theory says: recover the consumer's
read operator (\S\ref{sec:id}), water-fill bits on $\PC$ rather than on $I$
(\S\ref{sec:rd}), respect the channel's vacuity threshold (\S\ref{sec:capacity}), and
remember the verdict is relative to the observation budget (\S\ref{sec:rate}). Against
the literatures it unifies---indirect source coding, coding for computing,
rate--distortion--perception, task-oriented communication, loss-aware
compression---the contribution is making the task's metric a single identifiable
geometric object and testing it adversarially. The decisive validation is not
conceptual but empirical, and we ran it: a preregistered blind probe on an actual
Llama-3.2-3B attention layer (Gate~B, GO-P-2026-020). It \emph{misses} the sealed
bar. The real read operator is recovered at $4.5\times$ chance yet below the $0.60$
overlap threshold ($0.567$), and the registered---not reconstruction-matched---%
quantizer pair leaves the consumer-projected predictor unable to \emph{beat}
reconstruction (both call the winner on $16/16$ heads). The blind prediction direction
is right, but the decisive dissociation is untested on the trained model; on this
evidence the theory is bounded to synthetic consumers. The corrected experiment---%
reconstruction-matched key codes and a point-dependent $\PC(x)$ recovered per
activation---is the outstanding decisive test. Open
theory: we settle the fixed-$\PC$ achievability and converse of $R_{\mathcal{C}}(D)$
in Appendix~\ref{app:coding} and lift it to a fully point-dependent \emph{surrogate}
metric (Cor.~\ref{cor:pointdep}); what remains open is a single-letter coding theorem
for the \emph{true} nonlinear divergence at non-vanishing $D$ and a closed form for
the nonlinear tilt---no single tilted source; whether the flip is exactly
$R_{\mathcal{C}}$'s argmin migrating with $\PC$; and the consumer-relative
\emph{channel} coding theorem behind \S\ref{sec:capacity}. Underneath is one
principle: an object has no unique operational geometry---each observer
$\mathcal{O}=(\mathcal{C},G,B)$ pulls back one, and the quotient
$X/\!\sim_{\mathcal{C}}$ is the world it can act on. Four faces, one refutation,
eleven negatives: a world to map, now with coordinates in the literature.

\appendix
\section{The consumer-relative coding theorem}\label{app:coding}

Throughout the main theorem, $\PC$ is a \emph{fixed} PSD operator (constant, or
$\E_x[\PC(x)]$ under the proviso of Prop.~\ref{prop:main}); Cor.~\ref{cor:pointdep}
then lifts the operational statement to a point-dependent $\PC(x)$ surrogate metric.

\begin{lemma}[Reduction to the tilted source]\label{lem:tilt}
Let $Y=\PC^{1/2}X$, on $\operatorname{range}(\PC)$ of dimension $r=\operatorname{rank}\PC$.
Then $d_{\mathcal{C}}(x,\hat x)=\|y-\hat y\|^2$ with $\hat y=\PC^{1/2}\hat x$, and
$R_{\mathcal{C}}(D)=R_Y(D)$, the MSE rate--distortion function of $Y$; $\ker\PC$ costs
zero rate.
\end{lemma}
\begin{proof}
$d_{\mathcal{C}}=\|\PC^{1/2}(x-\hat x)\|^2=\|y-\hat y\|^2$. Objective and constraint
depend on $\hat X$ only through $\hat Y$, so by data processing $I(X;\hat X)\ge
I(Y;\hat Y)$; the minimizer places all information in $Y$ (a channel $p(\hat y\mid y)$,
then $\hat X=\PC^{-1/2}\hat Y$ on the range, conditional mean on $\ker\PC$), attaining
equality. Hence the minimum is $R_Y(D)$.
\end{proof}

\begin{theorem}[Achievability and converse]\label{thm:coding}
Let $X$ take values in a standard Borel space, $\{X_t\}$ i.i.d.\ $\sim p_X$, with a
measurable reconstruction alphabet and a reference letter $\hat x_0$ satisfying
$\E\,d_{\mathcal{C}}(X,\hat x_0)<\infty$ and $\E\,d_{\mathcal{C}}(X,\hat x_0)^2<\infty$.
For every $D>0$: \emph{(i)} for any $R>R_{\mathcal{C}}(D)$ there are $(2^{nR},n)$ codes
with $\limsup_n\E\,\tfrac1n\sum_t d_{\mathcal{C}}(X_t,\hat X_t)\le D$; \emph{(ii)}
every code of expected distortion $\le D$ has rate $\ge R_{\mathcal{C}}(D)$; and
$R_{\mathcal{C}}$ is convex and nonincreasing.
\end{theorem}
\begin{proof}
This is the single-letter rate--distortion theorem for a memoryless source on a
standard Borel space with a measurable difference distortion and finite
reference-letter moment \cite{berger1971}; we recall it in \emph{information-density}
form, which uses no differential entropies and so holds on general alphabets.
\emph{Achievability.} Fix $\varepsilon>0$ and a test channel $p^{*}(\hat x\mid x)$
with $\E\,d_{\mathcal{C}}\le D$ and $I(X;\hat X)\le R_{\mathcal{C}}(D)+\varepsilon$;
let $p_{\hat X}$ be the output marginal and $\imath(x;\hat x)=\log\frac{dP_{X\hat X}}%
{d(P_X\!\times\!P_{\hat X})}$ the information density. Draw
$\{\hat X^n(w)\}_{w=1}^{2^{nR}}$ i.i.d.\ $\sim\prod_t p_{\hat X}$; encode $x^n$ by any
$w$ with $\tfrac1n\sum_t\imath(x_t;\hat X_t(w))\le I(X;\hat X)+\varepsilon$
\emph{and} $\tfrac1n\sum_t d_{\mathcal{C}}(x_t,\hat X_t(w))\le D+\varepsilon$, else
declare failure and reconstruct with the constant $\hat x_0^{\,n}$. Let $\mathcal{T}$
be the joint event $\{\tfrac1n\sum_t\imath\le I+\varepsilon,\ \tfrac1n\sum_t
d_{\mathcal{C}}\le D+\varepsilon\}$; by the weak law $P_{X\hat X}^{n}(\mathcal{T})\to1$,
so for all but an $o(1)$-probability set $\mathcal{G}_n$ of source sequences the
\emph{conditional} probability $P_{\hat X^n\mid X^n=x^n}(\mathcal{T})\ge1-o(1)$.
Changing measure from $P_{\hat X^n\mid x^n}$ to the codebook marginal
$P_{\hat X^n}=\prod_t p_{\hat X}$ on $\mathcal{T}$ costs at most $2^{n(I+\varepsilon)}$
(there the density $\prod_t\frac{dP_{\hat X\mid X}}{dP_{\hat X}}=2^{\sum_t\imath}\le
2^{n(I+\varepsilon)}$), so for each $x^n\in\mathcal{G}_n$ a single random codeword
passes the encoding test with conditional probability $\ge2^{-n(I+2\varepsilon)}
(1-o(1))$. Hence $\Pr[\mathrm{fail}\mid x^n]\le\big(1-2^{-n(I+2\varepsilon)}
(1-o(1))\big)^{2^{nR}}\to0$ for $R>I+2\varepsilon$, uniformly over
$x^n\in\mathcal{G}_n$, and $\Pr[\mathrm{fail}]\le\Pr[\mathcal{G}_n^{c}]+
\sup_{x^n\in\mathcal{G}_n}\Pr[\mathrm{fail}\mid x^n]\to0$. On success the distortion is $\le D+\varepsilon$; on
failure it is $\tfrac1n\sum_t d_{\mathcal{C}}(X_t,\hat x_0)$, whose contribution to
the expectation is $\le\sqrt{\E[(\tfrac1n\sum_t d_{\mathcal{C}}(X_t,\hat x_0))^2]}\,
\sqrt{\Pr[\mathrm{fail}]}\to0$ by Cauchy--Schwarz, finite by the second-moment
hypothesis on $\hat x_0$. Thus $\limsup_n\E\,\tfrac1n\sum_t d_{\mathcal{C}}\le
D+\varepsilon$; let $\varepsilon\to0$.
\emph{Converse.} With $D_t=\E\,d_{\mathcal{C}}(X_t,\hat X_t)$,
$nR\ge H(W)\ge I(X^n;\hat X^n)\ge\sum_t I(X_t;\hat X_t)\ge\sum_t R_{\mathcal{C}}(D_t)
\ge nR_{\mathcal{C}}(\tfrac1n\sum_t D_t)\ge nR_{\mathcal{C}}(D)$, by memorylessness,
the definition of $R_{\mathcal{C}}$, Jensen (convexity), and monotonicity. Convexity
and monotonicity follow from time-sharing and distortion relaxation.
\end{proof}

\begin{corollary}\label{cor:coding}
$R_{\mathcal{C}}(D)$ equals the MSE rate--distortion of the $r$-dimensional tilted
source $Y=\PC^{1/2}X$; for Gaussian $X$ it is Prop.~\ref{prop:rd}; discretely,
$R_{\mathcal{C}}(0)=H(\Pi X)$ (Prop.~\ref{prop:floor}). The operational rate to serve
a consumer is governed entirely by the source projected into the read subspace---the
precise sense in which Shannon's theorem is the corner $\PC=I$, where the read
subspace is everything.
\end{corollary}

\begin{corollary}[Point-dependent surrogate]\label{cor:pointdep}
The coding theorem holds verbatim for the \emph{point-dependent} quadratic distortion
$d_{\PC}(x,\hat x)=(\hat x-x)^\top\PC(x)(\hat x-x)$ whenever $\PC(\cdot)$ is measurable
with $\|\PC(x)\|\le\Lambda<\infty$ and $\E\|X\|^4<\infty$. The achievability and
converse of Thm.~\ref{thm:coding} nowhere use that the distortion is a
\emph{difference} distortion---only that it is a nonnegative measurable single-letter
distortion with a finite second-moment reference letter, and $d_{\PC}(x,\hat x_0)\le
\Lambda\|\hat x_0-x\|^2$ supplies both. Hence the operational rate--distortion function
$R_{\mathcal{C}}(D)=\inf\{I(X;\hat X):\E\,d_{\PC}(X,\hat X)\le D\}$ exists and is
achievable for a fully point-dependent read operator.
\end{corollary}

\noindent What the fixed-$\PC$ specialization keeps and the point-dependent case
loses is therefore sharp: \emph{(i)} the closed-form reverse water-filling
(Cor.~\ref{cor:coding}) needs a single \emph{linear} tilt $Y=\PC^{1/2}X$, whereas with
$\PC(x)$ the tilt $x\mapsto\PC(x)^{1/2}x$ is nonlinear and $R_{\mathcal{C}}$ has no
closed form; \emph{(ii)} the exact floor $R_{\mathcal{C}}(0)=H(X/\!\sim_{\mathcal{C}})$
requires the integrability/compatibility conditions of \S\ref{sec:framework} to
identify the quotient; and \emph{(iii)} $d_{\PC}$ is the \emph{leading-order surrogate}
for the true divergence $D_Y(\mathcal{C}(x),\mathcal{C}(\hat x))$, so coding to it
certifies the true downstream distortion only in the high-rate (small-$D$) regime. The
genuinely open problem is thus not the existence of $R_{\mathcal{C}}(D)$ under a
point-dependent metric---Cor.~\ref{cor:pointdep} settles that---but a single-letter
coding theorem for the \emph{true} nonlinear divergence at non-vanishing $D$, together
with a closed form for the nonlinear tilt.

\begin{thebibliography}{99}
\bibitem{shannon1948} C.~E. Shannon. A mathematical theory of communication.
\emph{Bell Syst.\ Tech.\ J.}, 27:379--423, 623--656, 1948.
\bibitem{shannon1959} C.~E. Shannon. Coding theorems for a discrete source with a
fidelity criterion. \emph{IRE Nat.\ Conv.\ Rec.}, 4:142--163, 1959.
\bibitem{gauss1809} C.~F. Gauss. \emph{Theoria Motus Corporum Coelestium}, 1809.
\bibitem{berger1971} T.~Berger. \emph{Rate Distortion Theory}. Prentice-Hall, 1971.
\bibitem{dobrushin1962} R.~L. Dobrushin and B.~S. Tsybakov. Information transmission
with additional noise. \emph{IRE Trans.\ Inf.\ Theory}, 8(5):293--304, 1962.
\bibitem{wolfziv1970} J.~K. Wolf and J.~Ziv. Transmission of noisy information to a
noisy receiver with minimum distortion. \emph{IEEE Trans.\ Inf.\ Theory},
16(4):406--411, 1970.
\bibitem{witsenhausen1980} H.~S. Witsenhausen. Indirect rate distortion problems.
\emph{IEEE Trans.\ Inf.\ Theory}, 26(5):518--521, 1980.
\bibitem{korner1973} J.~K\"orner. Coding of an information source having ambiguous
alphabet and the entropy of graphs. \emph{Trans.\ 6th Prague Conf.}, 411--425, 1973.
\bibitem{orlitsky2001} A.~Orlitsky and J.~R. Roche. Coding for computing.
\emph{IEEE Trans.\ Inf.\ Theory}, 47(3):903--917, 2001.
\bibitem{blau2019} Y.~Blau and T.~Michaeli. Rethinking lossy compression: The
rate--distortion--perception tradeoff. \emph{ICML}, 2019.
\bibitem{gunduz2023} D.~G\"und\"uz et al. Beyond transmitting bits: Context,
semantics, and task-oriented communications. \emph{IEEE J.\ Sel.\ Areas Commun.},
41(1):5--41, 2023.
\bibitem{tishby1999} N.~Tishby, F.~C. Pereira, and W.~Bialek. The information
bottleneck method. \emph{Proc.\ 37th Allerton Conf.}, 368--377, 1999.
\bibitem{coifman2006} R.~R. Coifman and S.~Lafon. Diffusion maps. \emph{Appl.\
Comput.\ Harmon.\ Anal.}, 21(1):5--30, 2006.
\bibitem{calder2022} J.~Calder and N.~Garc\'ia Trillos. Improved spectral
convergence rates for graph Laplacians on $\varepsilon$-graphs and $k$-NN graphs.
\emph{Appl.\ Comput.\ Harmon.\ Anal.}, 60:123--175, 2022.
\bibitem{radovanovic2010} M.~Radovanovi\'c, A.~Nanopoulos, and M.~Ivanovi\'c. Hubs in
space. \emph{J.\ Mach.\ Learn.\ Res.}, 11:2487--2531, 2010.
\bibitem{lecun1990} Y.~LeCun, J.~S. Denker, and S.~A. Solla. Optimal Brain Damage.
\emph{NeurIPS}, 1990.
\bibitem{hassibi1993} B.~Hassibi and D.~G. Stork. Second order derivatives for
network pruning: Optimal Brain Surgeon. \emph{NeurIPS}, 1993.
\bibitem{dong2019hawq} Z.~Dong et al. HAWQ: Hessian aware quantization of neural
networks. \emph{ICCV}, 2019.
\bibitem{frantar2023gptq} E.~Frantar, S.~Ashkboos, T.~Hoefler, and D.~Alistarh.
GPTQ: Accurate post-training quantization for generative pre-trained transformers.
\emph{ICLR}, 2023.
\bibitem{paper1} A.~H. Bond. The angular observer: geodesic preservation and the
mode-budget inversion. Companion Paper~I, 2026.
\bibitem{paper2} A.~H. Bond. Geometric key quantization: reconstruction is not the
target. Companion Paper~II, 2026.
\bibitem{ledger2026} A.~H. Bond. \emph{Geometric Observation: claim ledger,
prediction registry, and reproduction artifacts}. Companion repository, 2026.
\end{thebibliography}

\paragraph{Reproducibility \& provenance.} All results resolve to committed registry
entries and artifacts in the public companion repository \cite{ledger2026}; each
registry entry is committed \emph{before} the run it governs, so the evidence that
registration preceded experimentation is the \emph{commit ordering} of the public
history. We are explicit about its limits: git and hosting timestamps are not
cryptographic third-party proof, and a snapshot taken \emph{after} completion
preserves the current history but does not retroactively establish timing. Stronger
anchoring---an external immutable deposit (Zenodo/OSF) and any contemporaneous
third-party timestamps (signed hashes, prior deposits)---accompanies the paper.
Computational experimentation and drafting were carried out with an AI research
assistant under the registration-first protocol of \S\ref{sec:negatives}; every
ledger row names its verifier and context-freshness.

\end{document}
