\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{array}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\newcommand{\PC}{P_{\mathcal{C}}}
\newcommand{\PCbar}{\bar P_{\mathcal{C}}}
\newcommand{\Sd}{\Sigma_{\delta}}
\newcommand{\Sx}{\Sigma_{x}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\dO}{d_{\mathcal{O}}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\kalign}{\kappa}
\newcommand{\seal}[1]{\texttt{GO-P-2026-#1}}

\title{\textbf{The Consumer-Relative Flip Across Domains}\\[2pt]
\large Preserve what the reader reads: a reconstruction--downstream dissociation
across acoustic, seismic, neural, and biological signals}
\author{Andrew H. Bond\\ \small \texttt{agi.hpc@gmail.com}}
\date{July 2026 --- \emph{Preprint (registration-first; every empirical claim resolves to a sealed ledger row).}}

\begin{document}
\maketitle

\begin{abstract}
Signals are almost universally compressed to minimise reconstruction error, on the
assumption that a more faithfully reconstructed signal better serves whatever reads it
next. For a consumer that reads only part of what a signal carries, this assumption is
directionally wrong: downstream error is governed not by the total reconstruction error
$\tr(\Sd)$ but by its projection onto the consumer's \emph{read operator},
$\tr(\PC\Sd)$. A code that spends a fixed budget minimising reconstruction spends it
largely on directions the consumer ignores; a code that instead protects the read
directions reconstructs the signal \emph{worse} yet serves the consumer \emph{better}
--- a dissociation we call the \textbf{flip}. We test the flip under a single
pre-registered protocol across twelve domains spanning three distinct physics:
direction finding on automotive radar, microphone, and seismic arrays; the attention
keys of a language model; and learned classifiers of moral judgment, music genre, and
sperm-whale dialect. The flip reproduces wherever the consumer's read operator is
\textbf{identifiable} --- given by the geometry, or recovered blind from the consumer by
probing what its output is sensitive to --- and fails, informatively, exactly where the
read operator is \emph{coupled} to the signal's own energy or the consumer does not
function. We derive the failure modes as the four degeneracies of $\PC$ and collapse the
central one into a continuous, ex-ante dial: a read--energy alignment coefficient
$\kalign=\tr(\PCbar\Sx)/\sum_{i\le r}\lambda_i(\Sx)$ whose endpoints are fixed
analytically (a coupling null at $\kalign\to1$; a maximal flip at $\kalign\to0$) and
whose interior is a registered prediction. Which code best exposes the flip is set not
by one statistic but by two axes of the representation's geometry --- concentration and
correlation --- a diagnostic that also selects the code at deployment and whose
single-statistic shortcut we pre-register and refute. Reconstruction fidelity is the
special case of a consumer that reads everything; observation, in general, is selective
--- and compression should be too.

\medskip
\noindent\textbf{Tiers, worn in public.} Seven sealed confirmations (across synthetic,
LLM-attention, acoustic, seismic, retrieval, and bioacoustic consumers) carry the claim;
one partial (radar, data-limited), one sealed null (gradient compression, the coupling
boundary), and two (A2)-probe verdicts (music, attention keys) are labelled as such
throughout. Every row resolves to a sealed pre-registration and a result artifact.
\end{abstract}

\section{Introduction}

Almost every system that compresses a signal is judged by how faithfully it can
reconstruct it --- mean-squared error for images, cosine similarity for embeddings, PSNR
for audio. The assumption underneath is so ingrained it is rarely stated: that a code
which reconstructs the signal better is a code that serves whatever reads the signal next
better. For a decoder that consumes only part of what the signal carries, this
assumption is not merely loose --- it is \emph{directionally wrong}.

The quantity that governs a consumer's downstream error is not the total reconstruction
error $\tr(\Sd)$ but its projection onto the consumer's \textbf{read operator},
$\tr(\PC\Sd)$ --- the error in the directions the consumer actually reads. When a fixed
bit budget is spent to minimise reconstruction, it is spent largely on directions the
consumer ignores; a code that instead protects the read directions reconstructs the raw
signal \emph{worse} yet serves the consumer \emph{better}. We call this dissociation the
\textbf{flip}: at matched bits, read-preserving coding beats reconstruction-optimal
coding on the downstream task, precisely while reconstruction-optimal coding wins on
reconstruction. The two fidelities move in opposite directions.

The flip was first isolated in two narrow settings --- the direction of arrival read by a
sensor array, and the keys read by softmax attention in a language model --- where it
could be dismissed as an artifact of the geometry. Here we ask whether it is instead a
\textbf{property of observation}. We test the same matched-bits dissociation, with the
same discipline (each configuration frozen on a calibration split and scored on held-out
data, every outcome pre-registered and reported regardless of sign), across twelve
domains spanning three distinct physics --- direction finding on automotive radar,
hearing-aid microphone arrays, and a seismic array reading earthquake back-azimuth; the
attention keys of a transformer; a fine-tuned moral-judgment classifier; music genre and,
strikingly, the dialect of sperm whales in their click-rhythm codas.

The dissociation reproduces across all three physics and every consumer type wherever a
single condition holds: the \textbf{read operator is identifiable} --- given by the
geometry, or recovered blind from the consumer by probing what its output is sensitive
to. Where it is not identifiable, or where it is \emph{coupled} to the signal's own
energy (as it is for gradient compression under a curvature metric, whose high-curvature
and high-energy directions coincide), the flip provably cannot appear --- and it does
not. These failures are not noise; they are the theory's scope conditions. We show they
are exactly the four ways the operator $\PC$ can degenerate, and that the central one
--- coupling --- is not a bucket but a \emph{continuous} coordinate, a read--energy
alignment coefficient $\kalign$ whose extremes we fix in closed form and whose interior
we register as a prediction. Which encoding best exposes the flip, in turn, lives on two
axes --- how \emph{concentrated} a representation's directions are and how
\emph{correlated} its channels are --- and no single summary statistic (we pre-register
one and refute it) can stand in for the full two-axis diagnostic.

\paragraph{Contributions.}
\begin{enumerate}\itemsep2pt
\item The flip demonstrated as \textbf{domain-general} --- seven sealed confirmations
spanning synthetic, LLM-attention, acoustic, seismic, retrieval, and bioacoustic
consumers, three distinct physics --- under sealed pre-registration with outcomes
reported regardless of sign.
\item A \textbf{blind read-operator probe} that recovers what a consumer reads without
its ground truth, turning a failed retrieval test into a confirmation on the identical
held-out set, and confirmed a second time on a fully virgin split --- the identifiability
boundary made constructive.
\item The failure modes derived as the \textbf{four degeneracies of $\PC$}, and the
coupling degeneracy collapsed into the \textbf{$\kalign$ alignment law} --- a continuous,
ex-ante predictor of flip magnitude with analytically fixed endpoints and a registered
interior point.
\item A \textbf{two-axis regime map} (concentration $\times$ correlation) that predicts
the winning code per domain, with a validated diagnostic and a refuted single-statistic
shortcut.
\end{enumerate}

\section{Theory}\label{sec:theory}

\subsection{The controlling quantity and the flip}

We take as given the second-order result established in the companion foundational work
(the COST paper): for a consumer $\mathcal{C}$ with output metric $G$ acting on a
representation through a Jacobian $J$, the leading-order downstream distortion induced by
a coding error $\delta$ is
\begin{equation}\label{eq:do}
\dO \;=\; \E_x \tr\!\big(\PC(x)\,\Sd(x)\big),\qquad \PC=\E[J^\top G J],
\end{equation}
reducing under stated conditions to $\tr(\PC\Sd)$, with $\Sd$ the error second moment.
$\PC$ is symmetric positive semidefinite; its range is the \emph{read subspace} (the
directions the consumer distinguishes) and its kernel is the quotient the consumer cannot
tell apart. Reconstruction fidelity is the special case $\PC=I$: a consumer that reads
everything.

Fix a total bit budget and consider three codes on the representation:
a \emph{read-preserving} code (R) that allocates bits to the range of $\PC$; a
\emph{reconstruction-optimal} code (O) that minimises $\tr(\Sd)$ (reverse water-filling
in the signal eigenbasis); and an \emph{anti} code (A) that starves the read directions.
The \textbf{flip} is the joint event
\begin{equation}\label{eq:flip}
\underbrace{d_{\mathcal C}(\text{R}) \;\le\; d_{\mathcal C}(\text{O})}_{\text{downstream: R wins}}
\quad\text{and}\quad
\underbrace{\tr(\Sd^{\text{O}}) \;\le\; \tr(\Sd^{\text{R}})}_{\text{reconstruction: O wins}} ,
\end{equation}
with the anti check $d_{\mathcal C}(\text{A}) \ge \max(d_{\mathcal C}(\text{R}),
d_{\mathcal C}(\text{O}))$. The two inequalities in \eqref{eq:flip} point in opposite
directions; that is the whole content. A code can win reconstruction and lose the task.

\subsection{The four degeneracies of $\PC$}\label{sec:taxonomy}

The flip in \eqref{eq:flip} is not automatic. It requires the reconstruction-optimal
allocation to \emph{differ} from the read-preserving one --- i.e.\ the read subspace must
be misaligned with the signal's high-energy directions --- and it requires the read
operator to be correctly known. There are exactly four ways this can fail, and each is a
degeneracy of $\PC$ (or of its estimate $\hat\PC$):

\begin{description}
\item[(D1) Coupling --- $\operatorname{range}\PC$ aligned with source energy.] If the
read subspace coincides with the top-energy eigenspace of $\Sx$, then O already allocates
bits to the read directions and R has nothing to add. The flip vanishes although $\PC$ is
exact. This is a genuine wall, not an artifact; it is the empirical content of the
gradient-compression null (\S\ref{sec:results}). We make it a continuous coordinate in
\S\ref{sec:kappa}.
\item[(D2) Precondition --- $\PC\approx0$.] If the consumer does not function, there is
no read subspace to protect and \eqref{eq:do} is degenerate. Observed for moral judgment
on frozen embeddings (chance accuracy): the flip is ill-posed until the consumer is
trained.
\item[(D3) Identifiability --- wrong $\hat G$ (hence wrong $\hat\PC$).] If the read
operator is \emph{estimated} rather than recovered from the consumer, the estimate can
protect the wrong subspace and O wins on held-out data. This is the only fixable failure:
recovering $\PC$ from the consumer's own output sensitivity (\S\ref{sec:probe}) restores
the flip.
\item[(D4) Mechanism-absence --- the claimed nuisance is not in $\ker\PC$.] If a
direction hypothesised to be invariant is not actually in the consumer's kernel,
protecting it buys nothing and the effect does not exist. A genuine refutation, not a
rehabilitatable miss.
\end{description}

\noindent A demonstration shows only the wins; a theory states where it fails. The four
degeneracies are what let us say, before any experiment, which domains admit the flip.

\subsection{The alignment law: coupling as a continuous dial}\label{sec:kappa}

Degeneracy (D1) is the informative one, because it is not binary. Let $\PCbar$ be the
orthogonal projector onto the rank-$r$ read subspace $\operatorname{range}\PC$, and let
$\lambda_1\ge\cdots\ge\lambda_d$ be the eigenvalues of the signal second moment $\Sx$.

\begin{definition}[read--energy alignment]\label{def:kappa}
\begin{equation}
\kalign \;=\; \frac{\tr(\PCbar\,\Sx)}{\sum_{i=1}^{r}\lambda_i(\Sx)}
\;\in\;\Big(\tfrac{\sum_{i>d-r}\lambda_i}{\sum_{i\le r}\lambda_i},\,1\Big].
\end{equation}
$\kalign$ is the fraction of the maximal $r$-dimensional signal energy that the read
subspace actually captures. $\kalign=1$ iff the read subspace \emph{is} the top-$r$
energy eigenspace of $\Sx$; $\kalign$ is small when the read subspace lies in the
signal's low-energy directions.
\end{definition}

\begin{proposition}[alignment law, informal]\label{prop:kappa}
At matched bits, the flip magnitude
$\Delta \;=\; d_{\mathcal C}(\mathrm{O}) - d_{\mathcal C}(\mathrm{R})$
is non-increasing in $\kalign$. At $\kalign=1$ (full coupling, D1) reverse water-filling
already spends its budget on the read directions, so $\Delta\to0$: the coupling null. As
$\kalign\to0$ the reconstruction-optimal code starves the read directions and $\Delta$ is
maximal.
\end{proposition}

\noindent The two endpoints are fixed \emph{analytically}, not fit:
\begin{itemize}\itemsep2pt
\item \textbf{$\kalign\to0$ (maximal flip).} For direction finding the read direction is
$\hat g = P_a^{\perp}a'(\theta_0)$, the array-manifold derivative orthogonalised against
the steering vector, which is by construction orthogonal to the dominant received energy
(the steering direction). For the synthetic and planted anchors the read subspace is
planted orthogonal to the signal energy. Both give $\kalign\approx0$ and the largest
observed flips.
\item \textbf{$\kalign\to1$ (null).} For gradient compression under the curvature metric
the read operator is the loss Hessian $H=\tfrac1n X^\top \mathrm{diag}(p(1{-}p))X$, whose
high-curvature directions coincide with the high-energy directions of $X^\top X$; hence
$\kalign\approx1$ and, as \S\ref{sec:results} confirms, $\Delta\approx0$ while the
anti-probe stays maximal --- the read operator governs the task, but the flip cannot
appear.
\end{itemize}

Proposition~\ref{prop:kappa} converts boundary condition (D1) from a bucket into an
ex-ante coordinate. Its endpoints are settled; its interior is the paper's registered
prediction. We measure $\kalign$ retrospectively across the twelve domains (labelled
retrospective) and register one prospective point at intermediate $\kalign$,
\seal{040}, predicting an intermediate $\Delta$ before the run
(Appendix~\ref{app:kappa}). \emph{This paper reports the analytic endpoints and the
registered protocol; the interior measurement is the one quantitative result still in
flight, and is marked as such wherever it appears.}

\subsection{Recovering the read operator}\label{sec:probe}

One rarely has $\PC$ in closed form. The read operator enters three ways.
\emph{Geometric} --- for sensor arrays it is $\hat g=P_a^{\perp}a'(\theta_0)$, known
analytically. \emph{Planted} --- in the synthetic and LLM-rematch anchors it is
constructed from a known subspace and the prediction made blind to it. \emph{Recovered}
--- on real learned representations it is obtained by the \textbf{blind probe}: perturb
the consumer's input and read its output sensitivity. For a cosine-ranking consumer the
margin sensitivity is $\partial\cos(a,b)/\partial b \approx \hat a-\cos(a,b)\,\hat b$
(the query component orthogonal to the item), and the read operator is the top-$r$
eigenspace of $S=\langle g\,g^\top\rangle$ --- low-rank for robustness, with $r$ and the
rate selected on an internal calibration split so that only a configuration that already
generalises once is sealed. The blind probe is the constructive form of the
identifiability face (D3): it is what makes the flip available on any real learned
representation whose consumer can be queried.

\section{Methods}\label{sec:methods}

\paragraph{A domain-general protocol.} Every domain is tested by the same design. On the
domain's native representation we build \textbf{three codes at matched total bits per
sample}: read-preserving (R), reconstruction-optimal (O, Lloyd--Max or reverse
water-filling), and anti (A). The consumer is the domain's own mature estimator,
independent of the compressor. For each held-out instance we record the downstream error
and the reconstruction error; the per-instance flip and anti checks are \eqref{eq:flip}.
All configuration is chosen on a calibration split and frozen before the held-out run;
each domain is pre-registered (a content hash sealed and git-timestamped before any
held-out instance is scored), and outcomes are recorded regardless of sign. Matched bits
are audited (allocation spread $\le1$--$2\%$); flip fractions are bootstrapped over
query subsets for the retrieval and classifier consumers.

\paragraph{The (A2) regime diagnostic.} For embedding consumers we additionally run
\texttt{probe\_quotient} (the compression engine's calibration-time probe): at matched
bits it applies a polar (angular), a per-channel (reconstruction-oriented), and a
ZCA-whitened proxy code and reports the Spearman rank agreement of the declared
consumer's scores under each. Two summary statistics characterise the regime:
\texttt{tangential\_fraction} ($\approx1\Rightarrow$ an angular consumer) and
\texttt{median\_unit\_displacement} ($\sqrt2\Rightarrow$ isotropic directions; small
$\Rightarrow$ a concentrated offset cone).

\paragraph{Domains and data.} \emph{Arrays} --- a simulated $\lambda/2$ ULA; LOCATA
(8\,cm nested ULA, OptiTrack ground truth); AV16.3 (8-mic circular array, camera-tracked
3-D mouth); PDAR (13-element IMS seismic array, USGS catalogue back-azimuth, 34
teleseismic events); RaDICaL (8-element 77\,GHz FMCW virtual array). \emph{Neural} ---
Llama-3.2-3B attention keys (layers 8/16); a logistic-regression training step (loss
Hessian as the read metric). \emph{Learned representations} --- CourtListener citation
pairs with LaBSE (cosine ranking); ETHICS commonsense with a bert-base classifier
fine-tuned to 0.74 (frozen embeddings sit at chance); FMA music with CLAP-PCA embeddings
and a genre classifier; sperm-whale codas (DSWP / Sharma 2024) with a clan/dialect
classifier on the inter-click-interval rhythm. Held-out sets are disjoint
recordings/events/opinions/codas wherever obtainable.

\section{Results}\label{sec:results}

Table~\ref{tab:sweep} is the spine; each row wears its tier.

\begin{table}[t]\small\centering
\caption{The twelve-domain sweep. Tier $\in$ \{sealed confirmation, partial, sealed
null, (A2) verdict\}. Every row resolves to a sealed pre-registration and a result
artifact (Appendix~\ref{app:domains}, Appendix~\ref{app:registry}).}
\label{tab:sweep}
\begin{tabular}{@{}llll@{}}
\toprule
Domain & Consumer / read operator & Held-out outcome & Tier \\
\midrule
Synthetic ULA & beamformer-ML + root-MUSIC / $\hat g$ & flip 6/6 (two estimators) & sealed conf. \\
LLM keys (Llama-3.2-3B) & softmax attention / blind probe & worse-arm 16/16, recon tied & sealed conf. \\
LOCATA acoustic & wideband MUSIC / array manifold & flip 11/13, recon-trade 13/13 & sealed conf. \\
AV16.3 acoustic & wideband MUSIC / array manifold & flip 74\%, recon-trade 100\% & sealed conf. \\
PDAR seismic & wideband MUSIC baz / array manifold & flip 76\%, recon-trade 100\% & sealed conf. \\
Legal retrieval & cosine ranking / blind probe & R 0.796 $>$ O 0.780 (virgin split) & sealed conf. \\
Whale codas & clan/dialect classifier / blind probe & R 0.934 $>$ O 0.883, flip 300/300 & sealed conf. \\
\midrule
RaDICaL radar & MUSIC DOA / virtual array & flip 25/25, anti 60\% (data-limited) & partial \\
Gradient optim. & curvature metric / loss Hessian & anti 300/300, flip 27\% & sealed null \\
\midrule
Music genre & genre classifier / weight & polar 0.994 $>$ per-ch 0.956 & (A2) verdict \\
Moral judgment & fine-tuned classifier / weight & polar preferred (once trained) & (A2) verdict \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{The flip reproduces across three physics.} On the physical arrays it holds
under the mature subspace estimator and against independent ground truth: LOCATA
\textbf{11/13} held-out recordings with the reconstruction-trade at \textbf{13/13} (Lloyd
reconstructs better every time yet is downstream-worse); AV16.3 \textbf{74\%} of held-out
windows with recon-trade \textbf{100\%} and anti-worst \textbf{76\%}, against
camera-tracked angle on \emph{disjoint} recordings; PDAR seismic back-azimuth
\textbf{76\%} of held-out events, recon-trade \textbf{100\%}, anti \textbf{76\%}, against
catalogue ground truth. The synthetic anchor flips 6/6 under two independent estimators;
the Llama-3.2-3B key consumer shows the read-operator-projected error predicting the
softmax-KL loser on \textbf{16/16} heads while exactly-tied reconstruction cannot
(a reconstruction-invisible \emph{worse-arm} discrimination, distinct from a two-consumer
inversion; \S\ref{sec:discussion}). The battery's pre-registered prediction --- a flip in
$\ge2$ of its three core prospective domains --- is met by the two confirmed arrays.

\paragraph{Recovering the read operator makes the flip transfer, twice.} On real
legal-citation retrieval an \emph{estimated} read operator (document covariance) overfits:
on held-out opinions reconstruction-optimal \emph{beats} read-preserving (AUROC 0.773 vs
0.757), a registered miss (\seal{035}, degeneracy D3). Replacing it with the blind-probe
read operator --- the only change --- reverses the verdict on the identical held-out set
(0.779 vs 0.771; flip 200/200 bootstraps) while reconstruction-optimal still reconstructs
2.6$\times$ better (\seal{036}). Because that held-out split had itself produced the 035
verdict and the margin was thin, we re-ran the \emph{frozen} recipe on a fully virgin,
opinion-disjoint split (evaluation opinions never previously scored), sealed before any
virgin instance was embedded (\seal{039}): all four bars pass and more cleanly ---
\textbf{R 0.796 $>$ O 0.780}, a margin double that of 036, with R edging even the
uncompressed embedding (0.794, nuisance-stripping denoises the ranking), recon
O 0.223 $\le$ R 0.561, flip 200/200, anti 200/200. Moral judgment shows the consumer
precondition (D2): frozen embeddings classify at chance (0.58--0.60 vs 0.63 majority) and
the flip is ill-posed; a classifier fine-tuned to 0.74 makes the angular code preserve the
class logit (0.994 vs 0.956 at 1 bit).

\paragraph{The whale flip is sealed, not exploratory.} Sperm-whale coda dialect is the
biologically most distant domain and the load-bearing cell of the regime map (isotropic
directions, yet correlated channels). It was promoted from an exploratory (A2) verdict to
a sealed flip (\seal{038}): with the clan classifier as consumer and the blind margin
probe as read operator, on held-out disjoint codas ($n=2907$) read-preserving AUROC
\textbf{0.934 $>$ 0.883} reconstruction-optimal, flip \textbf{300/300}, anti
\textbf{300/300}, while reconstruction-optimal reconstructs the click-interval features
2$\times$ better. The map's novel axis now rests on a confirmation.

\paragraph{Two boundary conditions bound the claim.} Gradient compression under a
curvature metric is the \emph{coupling} boundary (D1, $\kalign\approx1$): the anti-probe
is worst on \textbf{300/300} held-out steps --- the Hessian read operator genuinely
governs the update --- yet the flip sits at chance (\textbf{27\%}), because the
high-curvature and high-gradient-energy directions coincide, so reconstruction already
protects the read directions. No read-operator recovery can rescue it; the operator was
already exact. Radar is the data limit: the flip and reconstruction-trade hold
\textbf{25/25} on real 77\,GHz FMCW returns, but the anti control reaches only 60\% on the
8-element array and the disjoint-recording held-out is inaccessible (the full set is
310\,GB) --- recorded as a partial, not inflated to a confirmation.

\section{The regime map: two axes, not one}\label{sec:regime}

Characterising every embedding consumer with \texttt{probe\_quotient} places them on two
\emph{independent} axes --- direction concentration and cross-channel correlation --- and
the winning code is set by both (Table~\ref{tab:regime}). Music (isotropic, low
correlation) is preserved by the generic angular code and reconciles a pre-existing
$6\sigma$ music sign-flip; legal (concentrated, channel-structured) needs the
per-channel or blind-probe code; post-RoPE keys and sperm-whale codas both need the
whitened code, but for different reasons --- the keys are concentrated-correlated, the
codas isotropic-correlated (their inter-click intervals are temporally correlated). Whale
is the decisive case: isotropic like music yet whitening-dependent like the keys, so a
single concentration statistic cannot predict it. Indeed a pre-registered
\texttt{unit\_displacement} threshold is \emph{refuted} on its first out-of-sample test
(\seal{037}, ledger NEG-14): synthetic concentrated embeddings sit below the threshold
yet the generic polar code wins. The reliable diagnostic is the full end-to-end probe run
per domain at low bits.

\begin{table}[h]\small\centering
\caption{The two-axis regime map. Cells give the winning code; the empty cell
(concentrated / low-correlation) is argued structurally empty in
Appendix~\ref{app:domains}.}
\label{tab:regime}
\begin{tabular}{@{}l>{\centering\arraybackslash}p{3.2cm}>{\centering\arraybackslash}p{3.2cm}@{}}
\toprule
 & isotropic ($\text{unit\_disp}\approx\sqrt2$) & concentrated (unit\_disp small) \\
\midrule
low correlation & music, moral $\to$ \textbf{polar} & \emph{(structurally rare)} \\
cross-channel correlated & \textbf{whale $\to$ whitened} & keys $\to$ \textbf{whitened} \\
channel-structured & --- & legal $\to$ \textbf{per-channel / blind-probe} \\
\bottomrule
\end{tabular}
\end{table}

\section{Discussion}\label{sec:discussion}

\paragraph{What is established, and at what tier.} The load-bearing result is
\textbf{seven sealed confirmations} --- synthetic, LLM attention, LOCATA, AV16.3, PDAR,
legal (double-confirmed on a virgin split), and whale --- each scored on held-out data
against independent ground truth under a pre-registered protocol. Around them the sweep
carries, and labels, weaker evidence: one \textbf{partial} (radar, data-limited), one
\textbf{sealed null} (gradient compression), and two \textbf{(A2) verdicts} (music,
attention keys). Each wears its tier. We draw this line deliberately, because it is what
lets a reader trust the confirmations without re-running them: the same registration
machinery that seals a confirmation also makes an exploratory row legible as exploratory.
The full registry accounting --- every registration identifier, its disposition, and the
seal-before-run commit ordering --- is Appendix~\ref{app:registry}.

\paragraph{The flip and the (A2) verdict are not the same claim.} We reserve \emph{flip}
for the two-metric dissociation \eqref{eq:flip} established by the sealed protocol. The
\textbf{(A2) verdict} is a distinct, one-metric quantity: which quantizer family best
preserves the consumer's rank ordering at matched bits. A verdict is \emph{evidence
consistent with} the flip but does not by itself establish the reconstruction-trade. Two
rows report verdicts; seven report flips; the terms are kept apart throughout. (A separate
one-word disambiguation applies to the LLM-keys anchor: it is a reconstruction-invisible
\emph{worse-arm} discrimination on a single consumer, not a two-consumer verdict
\emph{inversion}.)

\paragraph{The failure taxonomy is what makes this a theory.} The four degeneracies of
$\PC$ (\S\ref{sec:taxonomy}) bound the claim where it fails. Identifiability is fixable by
recovery; coupling is a wall; precondition needs a working consumer; mechanism-absence is
a refutation. The gradient-compression null is coupling's empirical face, and the
$\kalign$ law (\S\ref{sec:kappa}) turns that wall into a graded, ex-ante prediction rather
than a category. The refuted single-statistic shortcut is the same discipline applied to
the regime map: a predictor pre-registered and killed on its first out-of-sample test,
which is what forces the regime to be characterised, not asserted.

\paragraph{The information-theoretic frame, and convergent evidence.} That reconstruction
fidelity is the special case $\PC=I$ places the flip inside rate--distortion theory as its
consumer-relative generalisation (the COST paper). Independent arrival at the same object
is telling: a bioacoustics decoder-robustness index --- the public name for the read-
operator-distortion measure whose provenance we footnote\footnote{Originally the ``Bond
Index'' in the ethics setting; renamed the Decoder Robustness Index (DRI) as the public
index name in the bioacoustic setting.} --- measures precisely read-operator distortion
(invariant acoustic transforms leave the decoder's read unchanged; stress transforms move
it), i.e.\ the same methodology reached from a different field.

\paragraph{Limitations.} The two reconciliations are (A2) verdicts, not sealed flips;
radar is data-limited; each domain uses a single model and dataset; the audio-level
decoder-robustness measurement was not run. The one quantitative piece still in flight is
the \emph{interior} of the $\kalign$ law: its endpoints are analytic and its protocol is
registered (\seal{040}), but the cross-domain $\kalign$ measurement is pending and is
marked as such wherever it appears --- we do not report $\kalign$ values we have not
measured.

\paragraph{Outlook.} The operational reading is direct: for compression under a known
consumer, \emph{measure the regime and adapt the code} --- certify what the consumer reads
rather than promise reconstruction. There is no universal code; there is a diagnostic, on
two axes, and a selector that reads it --- and, for the boundary, a single alignment
number that says in advance how much the flip will buy.

\appendix

\section{Per-domain detail}\label{app:domains}
Each row of Table~\ref{tab:sweep} expands to a per-domain entry recording the arms, the
sample size $n$, the sealed bars and the realised statistics with their binomial nulls,
the read-operator provenance (geometric / planted / recovered), and the seal and result
commit hashes. These reproduce the ledger's Gate-B rows and the statistical audit
(every reported percentage with its $n$, its bar, and its null); they are omitted here
for length and provided in the supplementary registry. The empty regime cell
(concentrated / low-correlation) is argued structurally rare: a representation whose
directions are concentrated into a shared offset cone is, generically, one whose channels
have acquired cross-channel correlation through that shared component --- so
concentration without correlation is the off-diagonal that the data do not populate.

\section{Registry accounting}\label{app:registry}
The ``no file drawer'' claim rests on the registration sequence having no unexplained
holes. The supplementary registry accounting gives every \texttt{GO-P-2026} identifier
(001--039) a disposition --- reported in this sweep, reported in a companion paper
(the COST theorem harnesses), superseded by a named successor (each miss still reported),
or never assigned (the two numbering skips, verified absent from the full git history) ---
together with the commit-ordering audit showing each sealed row's registration is a strict
git ancestor of its result. The supersession chains
($035\to036\to039$, $032\to033$, $020\to021$, and the earlier core iterations) carry every
miss at equal prominence with its correction.

\section{The $\kalign$ measurement protocol (\seal{040})}\label{app:kappa}
The retrospective analysis computes $\kalign$ (Definition~\ref{def:kappa}) for each of the
twelve domains from its read operator $\PC$ and signal second moment $\Sx$, and tests
Proposition~\ref{prop:kappa} (flip magnitude $\Delta$ non-increasing in $\kalign$) with the
analytic endpoints as anchors ($\kalign\approx0$ for the DOA/synthetic reads,
$\kalign\approx1$ for the gradient Hessian). The prospective seal \seal{040} fixes, before
the run, a domain engineered or selected to sit at intermediate $\kalign\approx0.5$ and
predicts an intermediate $\Delta$ bounded away from both the coupling null and the maximal
flip; the falsification is $\Delta$ landing at either extreme. This appendix is a protocol,
not a result: the $\kalign$ values are reported only once measured.

\end{document}
